References for PloS paper

### 2014-01-13 image_analysis_paper/plos_draft.tex
Added more to draft

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\usepackage{subcaption}
\usepackage{float}
\usepackage{pgffor}
%% END MACROS SECTION

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Image Analysis for {\em Drosophila} Embryogenesis}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carmeline J. Dsilva$^{1}$,
Bomyi Lim$^{1}$,
Stanislav Y. Shvartsman$^{1,2}$
Ioannis G. Kevrekidis$^{1,3,\ast}$
\\
\bf{1} Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, USA
\\
\bf{2} Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ, USA
\\
\bf{3} Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, USA
\\
$\ast$ E-mail: yannis@princeton.edu
\end{flushleft}

@@

The issue of rotation invariance arises in many applications, such as medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
%
In these settings, objects from different frames or timepoints must be aligned or ``registered'' before further analysis can be done.
%
%For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
%
%However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
%
Several different techniques have been developed to solve the image registration problem. 
%
``Template-based'' techniques are relatively common, where one selects a prototypical image or ``template'', and then aligns the remaining images to this template \cite{ahuja2007template}.
%
Such techniques can suffer if a good template is not know {\em a priori}; when the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments. 
%
Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, so that one can use these features (rather than the images themselves) in further analyses and computations.
%
However, constructing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and developing rotation-invariant features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
%
One can also construct algorithms that are invariant to rotations of the images \cite{hilai1994recognition, zhao2013fourier}.
%
ADD WHY THIS IS BAD.

We are interested in analyzing data collected from developing {\em Drosophila} embryos during nuclear cycle 14 (NC14) (CHECK).
%
(ADD MORE ABOUT WHY THIS IS IMPORTANT).

Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}; we can therefore image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI. 
%
We image the cross-section of each embryo.
%
We then use image analysis software (ADD HOW THIS IS DONE IN THE APPENDIX) to extract the circular dpERK concentration profile from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We measure the membrane thickness of them embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{...}. 
%
We then order the concentration profiles for each embryo by the thickness of the membrane to obtain a ``reconstruction'' of the dynamics of dpERK.

\begin{figure}[H]
\begin{center}
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.65\textwidth]{drosophila_imaging_setup}
\caption{}
\label{subfig:imaging_device}
\end{subfigure}
\begin{subfigure}{0.55\textwidth}
\includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
\includegraphics[width=0.3\textwidth]{drosophila_membrane}
\includegraphics[width=0.3\textwidth]{drosophila_dpERK}
\includegraphics[width=0.3\textwidth]{drosophila_DI}
\caption{}
\end{subfigure}
\begin{subfigure}{0.35\textwidth}
\centering
\includegraphics[width=0.6\textwidth]{circle_profile}\\
\includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
\caption{}
\end{subfigure}\\
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_ordered_membrane}
\caption{}
\end{subfigure}
\end{center}
\caption{{\bf Data collection.} (a) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
(b) Concentration profile of dpERK extracted from a fluorescent image. The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
(c) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(d) Concentration profiles of dpERK for many embryos from (c), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
\label{fig:background}
\end{figure}

Here,  we will discuss alternative methodologies to reconstruct the dpERK dynamics from fluorescent images. 
%
We will use various data mining techniques to {\em automatically} order the dpERK concentration profiles in time.
%
We will not only focus on the linear concentration profiles extracted from the images, but also work directly with the two-dimensional dpERK fluorescent images. 
%
Finally, we will look at ordering the two-dimensional membrane images in time; these are much more feature-rich than the dpERK images, and prove to be more challenging.

@@

We first used Principal Components Analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
%
We computed the first principal component, $\psi_1$, from the data set, and ordered the data by the projection coefficients onto $\psi_1$. 
%
The results are shown in Figure \ref{fig:PCA_ordering}.
%
There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness.

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\label{subfig:unordered_profiles}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_PCA}
\caption{}
\label{subfig:PCA_ordering_image}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{PCA_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using PCA.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(b) Concentration profiles of dpERK from (a), now ordered by the first PCA projection coefficient.
(c) Correlation between the first PCA projection coefficient and the membrane thickness.}
\label{fig:PCA_ordering}
\end{figure}

We would like to note that, because the embryos can be oriented with either the anterior or posterior end facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful. 
%
Therefore, we could also do PCA with an expanded data set where we also include the mirror image of each concentration profile.
%
Symmetries in PCA have been studied extensively in previous work \cite{holmes1998turbulence}; it was shown that when mirror images are also included in the data set, the first PCA mode captures the symmetric portion of the variability, and the second mode captures the anti-symmetric portion.
%
We found that including the mirror images of each data point did not change our results significantly, since the first principal component (not shown) is nearly symmetric with respect to left-right reflections.

@@

Although PCA was rather successful in ordering the concentration profiles, there are some inconsistencies in the orderings.
%
Perhaps the most obvious feature to note is the scrambling of the secondary *ADD NAME* peaks located at the far left and right edges of the concentration profiles.
%
It is known that these peaks arise later in the development of {\em Drosophila};
however, one can see in Figure \ref{subfig:PCA_ordering_image} that these secondary peaks are scrambled when the data is ordered using PCA.
%
We can gain insight into this ``scrambling'' by looking at the projection of the data onto the first two principal components, shown in Figure \ref{subfig:PCA_12}.
%
The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more approproate, as they can order the data along the curve, and not simply along the first principal axis.

We chose to use diffusion maps (DMAPS) as our nonlinear dimensionality reduction technique.
%
We computed the diffusion maps embedding for our data set, using the Euclidean distance between concentration profiles as our distance metric.
%
We then ordered the data using the first (nontrivial) diffusion maps coordinate $\phi_2$. 
%
The results are shown in Figure \ref{fig:DMAPS_ordering}.
%
The secondary peaks are now nicely clustered towards the end of the developmental trajectory.

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{coeff_12_new}
\caption{}
\label{subfig:PCA_12}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_DMAPS}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{DMAPS_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve. 
(b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
(c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
\label{fig:DMAPS_ordering}
\end{figure}

@@

As we have shown, we can order the concentration profiles using PCA or DMAPS.
%
However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
%
The developmental dynamics are invariant to the oritentation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
%
Rotation of the embryo cross-sections correspond to translation (with periodic boundary conditions) of the linear profiles we have been considering.
%
In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal protein peak.
%
However, without some {\em a priori} knowledge of the system (i.e. knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.

The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point or shape as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
%
Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy.
%
Angular synchronization \cite{singer2011angular} is a relatively new technique that is often more robust than template-based methods in the presence of noise.
%
We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
%
After aligning the data points using angular synchronization, we used DMAPS to order the aligned concentration profiles.
%
The results are shown in Figure \ref{fig:angsynch_ordering}.
%
One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}.

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_aligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_angsynch}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned using angular synchronization.
(c) Concentration profiles of dpERK aligned using angular synchronization and ordered by the first (non-trivial) DMAPS embedding coordinate.}
\label{fig:angsynch_ordering}
\end{figure}

@@

Instead of using angular synchronization and DMAPS, we can do the alignment and dimensionality reduction/ordering in one computation using vector diffusion maps (VDM).
%
VDM is similar to angular synchronization, but it is intended for data sets that contain both underlying symmetries and dynamics.
%
The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}).

\begin{figure}[H]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_vdm}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{VDM_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles using vector diffusion maps.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned and ordered using vector diffusion maps.
(c) Correlation between the VDM embedding coordinate and the membrane thickness.}
\label{fig:vdm_ordering}
\end{figure}

@@

Thus far, we have focused on the one-dimensional linear concentration profiles that are extracted from the fluorescent images.
%
However, the techniques we have outlined can also be applied to the two-dimensional fluorescent images. 
%
We use VDM to align and order the fluorescent images stained for dpERK. 
%
We now must factor out translations and rotations in the images. 
%
To do this, we project the images onto a sphere;
rotations and translations of a two-dimensional image then correspond to (three-dimensioanl) rotations of the sphere.
%
We then use VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
%
The results are shown in Figure \ref{fig:vdm_image_ordering};
the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.

\begin{figure}[H]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{vdm_2d_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in {1,9,17,25,30,34,38,43,49}{
\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_vdm_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the VDM embedding coordinate and the membrane thickness. 
(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. }
\label{fig:vdm_image_ordering}
\end{figure}

@@

An alternative to aligning the data (as in angular synchronization of VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between data points using these features for use in a DMAPS calculation.
%
In our setup, we need features for two-dimensional images that are invariant to translations and rotations of the images.
%
We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features.
%
Not only is this transform invariant to translations and rotations, but it is built on the wavelet transform (which is know to perform well for images), and is also contractive (so that noise and small deformations are not amplified by the transform).

We compute the scattering transform coefficients for each image in our data set.
%
We then perform a DMAPS calculation, using the Euclidean distance between the vectors of scattering transform coefficients as our distance metric.
%
The results are shown in Figure \ref{fig:scattrans_dpERK_ordering}.
%
The ordering is again good; however, using this method, we do not obtain optimal alignments for the images (we only obtain an ordering).

\begin{figure}[H]
\centering
\begin{subfigure}{\textwidth}
\includegraphics[width=0.7\textwidth]{DMAPS_scat_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in {1,9,17,25,30,34,38,43,49}{
\includegraphics[width=0.07\textwidth, trim=6in 0in 6in 0in, clip]{dpERK_scat_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness. The DMAPS embedding was computed using the translation- and rotation-invariant scattering transform coefficients as ``features'' of the images.
(b) Some representative fluorescent images, ordered using the first (non-trivial) DMAPS embedding coordinate computed from the scattering transform coefficients. }
\label{fig:scattrans_dpERK_ordering}
\end{figure}

>>>>

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW
\usepackage{subcaption}
\usepackage{float}
\usepackage{pgffor}
\usepackage{hyperref}

%% END MACROS SECTION

@@

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Rotation-Invariant Image Analysis for Reconstructing Dynamics in {\em Drosophila} Embryogenesis}
}
% Insert Author names, affiliations and corresponding author email.
\\
Carmeline J. Dsilva$^{1}$,
Bomyi Lim$^{1}$,
Stanislav Y. Shvartsman$^{1,2}$
Ioannis G. Kevrekidis$^{1,3,\ast}$
\\
\bf{1} Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, USA
\\
\bf{2} Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, NJ, USA
\\
\bf{3} Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, USA
\\
$\ast$ E-mail: yannis@princeton.edu
\end{flushleft}

@@

The issue of rotation invariance arises in many applications, including medical imaging \cite{hajnal2010medical}, face recognition \cite{rowley1998rotation}, digit recognition \cite{simard1992efficient}, and texture classification \cite{greenspan1994rotation}.
%
In these settings, objects from different frames or timepoints must be aligned or ``registered'' prior to further analysis.
%
%For example, often medical images (such as MRI scans) are collected over multiple days, and doctors would like to study development of a condition (such as the growth of a tumor).
%
%However, on each day, the patient and/or sensors may be in slightly different locations; therefore, the images must be aligned before the doctor can look at how the tumor is growing and evolving.
%
Several different techniques have been developed to solve the image registration problem. 
%
``Template-based'' techniques, where one selects a prototypical image or ``template'' and then aligns the remaining images to this template, are relatively common \cite{ahuja2007template}.
%
Such techniques can suffer if a good template is not know {\em a priori}; if the images are noisy, selecting one image from the data set at random can yield a noisy template and result in many mis-alignments. 
%
Other techniques focus on constructing ``features'' that are rotation-invariant \cite{flusser2000independence, lowe1999object, sadler1992shift, ojala2002multiresolution}, so that one can use these features (rather than the images themselves) in further analyses and computations.
%
However, developing rotation-invariant features that still retain the important information in the images is difficult and often problem-dependent, and constructing features that are stable to noise and deformations is still an active area of research \cite{mallat2012group, sifre2013rotation}.
%
One can also construct algorithms that consider all possible image rotations in the analysis \cite{hilai1994recognition, zhao2013fourier}.
%
%TODO: why do these methods fail?

We are interested in analyzing images of developing {\em Drosophila} embryos during nuclear cycle 14 (NC14).
%TODO: check that NC14 is correct
%
More specifically, we would like to order the images {\em in time} so that we can reconstruct the developmental dynamics from snapshots. 
%
% TODO: add motivation about why we are interested in this
%
Currently, data is collected using a high-throughput microfluidic device \cite{chung2010microfluidic}, so we can image hundreds of embryos in one experiment.
%
Each embryo is stained for the membrane proteins, dpERK, and DI, and we collect fluorescent images of the cross-section of each embryo using a confocal microscope.
%
We then use image analysis software to extract the dpERK concentration profile around the circumfrence of the embryo from the fluorescent image.
%
We use the expression of DI to find the dorsal midline of the embryo, and ``unzip'' the circular profile at the dorsal midline to obtain a one-dimensional, linear dpERK concentration profile.
%
We also measure the membrane thickness of the embryo; the thickness of the membrane is known to be monotonic in time during NC14 \cite{lim2013kinetics, lecuit2002slam}. 
%
We then order the dpERK concentration profiles for each embryo by the thickness of the membrane to obtain a ``trajectory'' of the dpERK dynamics.

\begin{figure}[!ht]
\begin{center}
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.65\textwidth]{drosophila_imaging_setup}
\caption{}
\label{subfig:imaging_device}
\end{subfigure}
\begin{subfigure}{0.55\textwidth}
\includegraphics[width=0.8\textwidth]{drosophila_schematic}\\
\includegraphics[width=0.25\textwidth]{drosophila_membrane}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_dpERK}
\hfill
\includegraphics[width=0.25\textwidth]{drosophila_DI}
\caption{}
\end{subfigure}
\begin{subfigure}{0.35\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{circle_profile}\\
\includegraphics[width=\textwidth, trim=0mm 300mm 0mm 200mm, clip]{line_profile}
\caption{}
\end{subfigure}\\
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\includegraphics[width=\textwidth]{data_ordered_membrane}
\caption{}
\end{subfigure}
\end{center}
\caption{{\bf Data collection.} (a) Schematic of {\em Drosophila} embryo (longitudinal and cross-section), and cross-sectional fluorescent images of an embryo stained for the membrane proteins (left), dpERK (center), and dorsal protein (right).
(b) Concentration profile of dpERK extracted from a fluorescent image. The circular profile is ``unwrapped'' at the dorsal midline (determined from the expression of the dorsal protein) to obtain a concentration profile on a line.
(c) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(d) Concentration profiles of dpERK for many embryos from (c), now ordered by the thickness of the membrane. The membrane thickness is known to be one-to-one with time.}
\label{fig:background}
\end{figure}

Here,  we will discuss alternative methodologies to reconstruct the dpERK dynamics from fluorescent images. 
%
We will use various dimensionality techniques to {\em automatically} order the dpERK concentration profiles in time.
%
We assume that the data is approximately one-dimensional, and that this one dimension parameterizes time, so that reducing the data to one-dimension and ordering it along this dimension will (approximately) order it in time. 
%
We will not only focus on the linear concentration profiles extracted from the images, but also work directly with the two-dimensional dpERK fluorescent images. 
%
Finally, we will look at ordering the two-dimensional membrane images in time; these are much more feature-rich than the dpERK images, and prove to be more challenging.

@@

We first used principal components analysis (PCA) to automatically order the one-dimensional concentration profiles in time. 
%
We computed the first principal component, $\psi_1$, from the data set, and ordered the data by the projection coefficients onto $\psi_1$.
%
The results are shown in Figure \ref{fig:PCA_ordering}.
%
There is a strong correlation between the first projection coefficient, $\langle x_i, \psi_1 \rangle$, and the membrane thickness;
the Spearman correlation between the first projection coefficient and the membrane thickness is 0.9244.

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unordered}
\caption{}
\label{subfig:unordered_profiles}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_PCA}
\caption{}
\label{subfig:PCA_ordering_image}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{PCA_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using PCA.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time.
(b) Concentration profiles of dpERK from (a), now ordered by the first PCA projection coefficient.
(c) Correlation between the first PCA projection coefficient and the membrane thickness.}
\label{fig:PCA_ordering}
\end{figure}

We would like to note that, because the embryos can be oriented with either the anterior or posterior facing up in the microfluidic device (see Figure \ref{subfig:imaging_device}), the directionality of the concentration profiles (clockwise versus counterclockwise around the ring, or right-to-left versus left-to-right for the linear profiles) is not meaningful. 
%
Therefore, we could do PCA with an expanded data set where we also include the mirror image of each concentration profile.
%
Symmetries in PCA have been studied extensively in previous work \cite{holmes1998turbulence}; it was shown that when mirror images are also included in the data set, the first PCA mode captures the symmetric portion of the variability, and the second mode captures the anti-symmetric portion.
%
We found that including the mirror images of each data point did not change our results significantly, since the first principal component (not shown) is nearly symmetric with respect to left-right reflections.

@@

Although PCA was rather successful in ordering the concentration profiles, there are some inconsistencies in the orderings.
%
Perhaps the most obvious feature to note is the scrambling of the secondary peaks located at the far left and right edges of the concentration profiles.
%TODO: add name of secondary peaks
%
It is known that these peaks arise later in the development of {\em Drosophila} \cite{...};
however, one can see in Figure \ref{subfig:PCA_ordering_image} that these secondary peaks are scrambled when the data is ordered using PCA.
%
We can gain insight into this ``scrambling'' by looking at the projection of the data onto the first two principal components, shown in Figure \ref{subfig:PCA_12}.
%
The data falls onto a one-dimensional {\em nonlinear} curve in this two-dimensional space, implying that nonlinear dimensionality reduction techniques may be more appropriate.

We chose to use diffusion maps (DMAPS) as our nonlinear dimensionality reduction technique.
%
We computed the diffusion maps embedding for our data set, using the Euclidean distance between concentration profiles as our distance metric.
%
We then ordered the data using the first (nontrivial) diffusion maps coordinate $\phi_2$. 
%
The results are shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane thickness is 0.9209.
%
The secondary peaks are now nicely clustered towards the end of the developmental trajectory.

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{coeff_12_new}
\caption{}
\label{subfig:PCA_12}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_DMAPS}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{DMAPS_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering dpERK concentration profiles using DMAPS.} (a) Projection of data shown in Figure \ref{subfig:unordered_profiles} onto the first two principal components; note that the data falls on an (effectively) one-dimensional, nonlinear curve. 
(b) Concentration profiles of dpERK from (a), now ordered by the first non-trivial DMAPS embedding coordinate.
(c) Correlation between the first (non-trivial) DMAPS embedding coordinate and the membrane thickness.}
\label{fig:DMAPS_ordering}
\end{figure}

@@

We have demonstrated that we can order the concentration profiles using PCA or DMAPS.
%
However, ordering the profiles first requires that we factor out the relevant symmetries in the problem.
%
The developmental dynamics are invariant to the orientation of the embryo in the microfluidic device; therefore, to obtain meaningful results from our analysis, we must first align the embryo snapshots.
%
Rotations of the embryo cross-sections correspond to translations (with periodic boundary conditions) of the linear profiles we have been studying.
%
In the previous analyses, we have assumed that the profiles have been prealigned using the location of the dorsal protein peak.
%
However, without some {\em a priori} knowledge of the system (e.g. without knowing that dorsal is expressed at one position along the perimeter of the embryo), we would need to align the concentration profiles before doing our analysis.
%TODO: add when this might be relevant, e.g. when there is no Dorsal

The most common and perhaps intuitive techniques for alignment are template-based methods \cite{ahuja2007template}, where one chooses are representative data point or shape as a ``template'', and then shifts or rotates each data point to be optimally aligned with this template.
%
Although these methods are simple, they require knowledge of an appropriate template function and can suffer when the data set is noisy \cite{sonday2013noisy}.
%
Angular synchronization \cite{singer2011angular} is a relatively new technique that is often more robust than template-based methods in the presence of noise.
%
We used angular synchronization to factor out the shifts in the concentration profiles measured from the fluorescent images.
%
After aligning the data points using angular synchronization, we used DMAPS to order the aligned concentration profiles.
%
The results are shown in Figure \ref{fig:angsynch_ordering}.
%
One can see that the results are visually consistent with the results shown in Figure \ref{fig:DMAPS_ordering}; the Spearman correlation between $\phi_2$ and the membrane length is 0.9164.

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_aligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_angsynch}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned using angular synchronization.
(c) Concentration profiles of dpERK aligned using angular synchronization and ordered by the first (non-trivial) DMAPS embedding coordinate.}
\label{fig:angsynch_ordering}
\end{figure}

@@

Instead of using angular synchronization followed by DMAPS, we can do the alignment and ordering in one computation using vector diffusion maps (VDM).
%
VDM is similar to angular synchronization, but it operates on data sets that contain both underlying symmetries and dynamics.
%
The results of aligning {\em and} ordering the data using VDM are shown in Figure \ref{fig:vdm_ordering}; the results are similar to those from angular syncrhonization (Figure \ref{fig:angsynch_ordering}), and the Spearman correlation between $\langle \phi_3, \phi_1 \rangle$ and the membrane length is 0.9123.

\begin{figure}[!ht]
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_unaligned_unordered}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{data_ordered_vdm}
\caption{}
\end{subfigure}
\begin{subfigure}{0.3\textwidth}
\includegraphics[width=\textwidth]{VDM_time_corr}
\caption{}
\end{subfigure}
\caption{{\bf Ordering spatially unaligned dpERK concentration profiles using vector diffusion maps.} (a) Concentration profiles of dpERK for many embryos. Each row represents a different embryo fixed at a slightly different developmental time. The circluar profiles have not been ``opened'' at the same point; rather, each profile was opened at a random point around the embryo. We are allowed to shift the (linear) profiles left and right.
(b) Concentration profiles of dpERK aligned and ordered using vector diffusion maps.
(c) Correlation between the VDM embedding coordinate and the membrane thickness.}
\label{fig:vdm_ordering}
\end{figure}

@@

Thus far, we have focused on the one-dimensional linear concentration profiles that are extracted from the fluorescent images.
%
However, the techniques we have outlined can also be applied to the two-dimensional fluorescent images. 
%
We used VDM to align and order the dpERK fluorescent images. 
%
We now must factor out translations and rotations in the images. 
%
We first projected the images onto a sphere;
rotations and translations of a two-dimensional image then correspond to (three-dimensional) rotations of the sphere.
%
We then used VDM to align and order the (projected) images, with the underlying symmetry group $SO(3)$. 
%
The results are shown in Figure \ref{fig:vdm_image_ordering};
the orderings are of comparable quality to those obtained with the one-dimensional profiles, and the dpERK peaks are nicely aligned.
%
The Spearman correlation between $\langle \phi_4, \phi_1 \rangle$ and the membrane length is 0.9051.

\def\imageindices{1,19,25,30,35,40,45,49}
%\def\imageindices{1,9,17,25,30,34,38,43,49}

\begin{figure}[!ht]
\centering
\begin{subfigure}{\textwidth}
\centering
\includegraphics[width=0.7\textwidth]{vdm_2d_time_corr}
\caption{}
\end{subfigure}
\begin{subfigure}{\textwidth}
\foreach \n in \imageindices{
\includegraphics[width=0.1\textwidth]{dpERK_vdm_\n}
\hfill}
\caption{}
\end{subfigure}
\caption{{\bf Ordering two-dimensional fluorescent images of dpERK.}
(a) Correlation between the VDM embedding coordinate and the membrane thickness. 
(b) Some representative fluorescent images, aligned  and ordered using the VDM embedding coordinate. The corresponding data points are circled in (a). }
\label{fig:vdm_image_ordering}
\end{figure}

@@

An alternative to aligning the data (as in angular synchronization or VDM) is to construct {\em features} that are invariant to the underlying symmetry group, and then compute distances between features of the data for use in a DMAPS calculation.
%
In our setup, we require features of two-dimensional images that are invariant to translations and rotations of the images.
%
We chose to use the scattering transform \cite{mallat2012group} to construct our invariant features; the scattering transform utilizes a series of wavelet transforms, which are known to perform well for images \cite{akansu2010emerging}.
%
Not only is this transform invariant to translations and rotations, and is also contractive (so that noise and small deformations are not amplified by the transform) \cite{bruna2011classification}.

We compute the scattering transform coefficients for each image in our data set.
%
We then perform a DMAPS calculation, using the Euclidean distance between vectors of scattering transform coefficients as our distance metric.
%
The results are shown in Figure \ref{fig:scattrans_dpERK_ordering}.
%
The ordering is again good, with a Spearman correlation between $\phi_2$ and the membrane length of 0.9011.
%
However, we would like to note that using this method, we do not obtain optimal alignments for the images (we only obtain an ordering).
